### 2025-02-19 paper/sections/method.qmd

Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$. Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\mathbf{y}^{+}$ denotes some target output. We say that $\mathbf{M}_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are plausible (i.e. consistent with the data) and actionable. Formally, we define these properties as follows:

1. (Plausibility) $\int^{A} p(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
2. (Actionability) Permutations $\Delta$ are subject to actionability constraints.

@@

3. (Faithfulness) $\int^{A} p_\theta(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.

@@

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote iteratively generated counterfactuals. To train models with high explainability as defined in @def-explainability, we propose to leverage counterfactuals in the following objective,

@@

We have simulated this scenario using synthetic data with independent features and Gaussian class-conditional densities in @fig-poc. The four panels in @fig-poc show the outcomes for different training procedures using the same model architecture each time (a linear classifier). In each case, we show the linear decision boundary (green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$. Stars indicate counterfactuals in the target class generated at test time using generic gradient descent for a fixed number of iterations.

>>>>

Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$. Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.

We say that $\mathbf{M}_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are plausible (i.e. consistent with the data) and actionable. Formally, we define these properties as follows:

1. (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
2. (Actionability) Permutations $\Delta$ are subject to actionability constraints.

@@

3. (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.

@@

Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual. For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.].

To train models with high explainability as defined in @def-explainability, we propose to leverage counterfactuals in the following objective,

@@

We have simulated this scenario using synthetic data with independent features and Gaussian class-conditional densities in @fig-poc. The four panels in @fig-poc show the outcomes for different training procedures using the same model architecture each time (a linear classifier). In each case, we show the linear decision boundary (green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$. Stars indicate counterfactuals in the target class generated at test time using generic gradient descent until convergence.

### 2025-02-19 paper/sections/other/counterfactuals.qmd

An important consideration when generating counterfactual explanations using gradient-based methods is how to define convergence.

>>>>

An important consideration when generating counterfactual explanations using gradient-based methods is how to define convergence. Two common choices are to 1) perform gradient descent over a fixed number of iterations $T$, or 2) conclude the search as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$. We prefer the latter for our purposes, because it explicitly defines convergence in terms of the black-box model, $\mathbf{M}(\mathbf{x})$.

Defining convergence in this way allows for a more intuitive interpretation of the resulting counterfactual outcomes than with fixed $T$. Specifically, it allows us to think of counterfactuals as explaining 'high-confidence' predictions by the model for the target class $y^+$. Depending on the context and application, different choices of $\tau$ can be considered as representing 'high-confidence' predictions.
